---
title: "Matrices"
output: 
  html_document:
    toc: true
    toc_float: true
    collapsed: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(igraph)
library(Matrix)
library(miniCRAN)
```

## Dense Matrices
A matrix is a structure in `R` with two dimensions. Rows and columns can have names which can be changed without changing the values of the matrx.

Obtaining the row of a matrix gives the result as a matrix, adding `drop = F` avoids this.

R also has higher-dimensional structures called arrays.

### Solving linear systems
To solve $Ax = b$ we can write it in two ways. One is to invert $A$ first and multiply by $b$ (`solve(A) %*% b`). The other is to use `solve` directly to solve for $x$. Finding the inverse first results in less accuracy and is slower so it should not be done unless the same system needs to be solved for many different vectors.

```{r }
A = matrix(c(rnorm(10^2, 3, 0.5)), nrow = 10)
b = rnorm(10)
norm(A %*% solve(A, b) - b)
norm(A %*% solve(A)%*% b - b)
```

In this case we see the numerical error is slightly larger but this generally increases with the dimension of the matrix.

### Numerical stability
In `R` a floating point number is stored as a `double` precision number. According to `IEEE754` standard a double number has 64 binary bits with:

* sign bit ($s$): 1 bit

* exponent ($e$): 11 bits

* significant precision ($m$): 52 bits

The decimal equivalent is $(-1)^s \times 2^{e - 1023} \times 1.m$. We expect an error of order $10^{-16}$ in numerical representations of double numbers. A simple example to demonstrate this is:

```{r numtest}
0.1 + 0.2 - 0.3
```

To account for numerical innaccuracy we can use `all.equal` instead. These errors compound within matrix multiplication so can often have larger errors. `Solve(A,b)` incurs less numerical error than `solve(A)`. Eigenvalues can also involve high numerical error, if you know an input is symmetric this should be specified in the input argument.

## Sparse Matrices
Sparse and dense matrices can be accessed with the `Matrix` package. Dense matrices are stored as `dgeMatrix`. It adds the `rankMatrix` and `rcond` functions. A sparse matrix is one where most entries are 0. It is inefficient to store all these 0's in the memory. Sparse matrices are generally stored as `dgCMatrix`. d stands for digit, g for general (not triangular or symmetric) and C for column (not row or triplet). A sparse matrix can be converted to dense at the cost of increased memory storage. A `dgCMatrix` is stored using the compressed sparse column (CSC) format. This stores the length of runs of zeros and non-zero entries. 

```{r sparse}
library(Matrix)
set.seed(0)
nrows = 100; ncols = 100
vals = sample(x=c(0, 2, 4, 9), prob=c(0.97, 0.01, 0.01, 0.01), size=nrows*ncols, replace=TRUE)
m1 = matrix(vals, nrow=nrows, ncol=ncols)
m2 = Matrix(vals, nrow=nrows, ncol=ncols, sparse = TRUE)
m1[1:2, 1:10]
c(object.size(m1), object.size(m2))
```

We can see the sparse format is far more efficient with memory usage.

### Basic operations
Multiplication and division retain the `dgCMatrix` format but addition or subtraction turn `dgC` into `dge`. A sparse matrix multiplied by a dense vector outputs a `dge` but two sparse matrices can be multiplied together and retain sparsity.

### Obtaining the dependency graph of R packages.
We can use the `miniCRAN` package to build an `R` package dependency graph. This can be represented via an adjacency matrix `M`. The matrix is square with each row and column relating to a vertex. If there is an edge from $i$ to $j$ then $M_{i,j} =1$ else the entry is 0. Thus $M$ is usually sparse. We look at the dependencies of `rmarkdown` below: 

```{r package}
library(miniCRAN) # for obtaining dependency relations of R packages
g1 <- miniCRAN::makeDepGraph('rmarkdown', suggests = FALSE, enhances = FALSE)
plot(g1)
get.adjacency(g1)
```

So clearly this adjacency matrix would be a good candidate for a sparse representation.

### Solving large linear systems
Sparse matrices can be even worse when inverting, the inverse of a sparse matrix is not necessarily sparse, thus we should use `solve(A,b)` whenever possible. 